---
title: "Pretty ggplots with custom themes, ggtext, and ggh4x"
output:
  md_document:
    variant: gfm
preserve_yaml: true
toc: true
---

(delete this)`---
toc: true
title: "Pretty ggplots with custom themes, ggtext, and ggh4x"
excerpt: "A quick dissection of a plot I'm pretty proud of."
---`(delete this)

```{r opts, echo = FALSE, warning=F, message=F}
title <- "ggplot-extensions-and-custom-themes"
knitr::opts_knit$set(base.dir = "/Users/MJ/Github/mcmaurer.github.io/", base.url = "/")
knitr::opts_chunk$set(fig.path = paste0("assets/rmd-images/", title, "/"))
library(tidyverse)
library(MCMsBasics)
library(ggh4x)
library(ggtext)
```


# Intro

Sometimes I have an idea of something I'd like to share that's too big for a tweet thread or entry in my [R Snippets repo](https://github.com/MCMaurer/R_snippets), but I'm too lazy to flesh it out into a full, nicely written blog post. I find these ideas are often useful, but end up floundering and remain unshared. To counteract that, I'm just gonna do a rough sketch blog post walking through some useful bits of a recent exercise of mine.

I have been working on a new `ggplot2` theme for my own personal use, not for any professional settings. I wanted something fun and distinctive, with flexibility to push plots towards a more polished and completed look. While working on building the theme, I found myself pushing one of my example plots into one of these more polished, cool-looking plots you sometimes see shared online. Now, mine is paltry compared to some of the really cool data visualizations people make in R. 

There are a few recent blog posts and visualizations I've seen that inspired me to try making some more in-depth plots. Albert Rapp has a few fantastic blog posts recreating more complex plots [here](https://albert-rapp.de/posts/ggplot2-tips/11_rounded_rectangles/11_rounded_rectangles.html) and [here](https://albert-rapp.de/posts/ggplot2-tips/10_recreating_swd_look/10_recreating_swd_look.html), and Bob Rudis does something similar [here](https://rud.is/b/2022/01/04/starting-2022-off-with-a-fairly-complex-ggplot2-recreation-plot/). Finally, the [filmicaesthetic GitHub account](https://github.com/filmicaesthetic/Art-and-Music) has a couple gorgeous plot/infographics to check out.

Here is my final plot and the code necessary to generate it:

```{r initial-plot, fig.height=8, fig.width=10}
p <- diamonds %>% 
  mutate(cutname = "Cut", clarityname = "Clarity",
         color = fct_rev(color), 
         clarity = fct_rev(clarity)) %>% 
  ggplot(aes(x = carat, y = price, color = color)) +
  geom_point(size = 0.2, alpha = 0.8) +
  theme_mcm_dark(facet_outlines = T) +
  scale_color_viridis_d(option = "A", begin = 0.25, 
                        guide = guide_stringlegend(nrow = 1, title.vjust = 1.1)) +
  theme(legend.position = "top",
        legend.margin = margin(0,0,0,0),
        legend.box.margin = margin(10,0,-20,0),
        legend.direction = "horizontal",
        legend.justification = "left",
        axis.ticks = element_line(), 
        axis.title.x = element_text(hjust = 0.065),
        axis.title.y = element_text(hjust = 0.015)) +
  facet_nested(rows = vars(clarityname, clarity), 
               cols = vars(cutname, cut),
               strip = strip_nested(text_x = 
                                      elem_list_text(
                                        family =
                                        c("FrauncesSuperSoftWonky-Medium",
                                          rep(NA, 100))),
                                    text_y = 
                                      elem_list_text(
                                        family =
                                        c("FrauncesSuperSoftWonky-Medium", 
                                            rep(NA, 100)))),
               scales = "free") +
  facetted_pos_scales(
    x = list(
      cut == "Fair" ~ scale_x_log10(limits = range(diamonds$carat)),
      cut != "Fair" ~ scale_x_log10(limits = range(diamonds$carat), 
                                    labels = NULL)
    ),
    y = list(
      clarity == "I1" ~ scale_y_continuous(limits = range(diamonds$price), 
                                           labels = scales::label_dollar()),
      cut != "I1" ~ scale_y_continuous(limits = range(diamonds$price), 
                                         labels = NULL)
    )
  ) +
  labs(title = "Diamond prices are affected by multiple variables",
       subtitle = "Size matters, but so do cut, clarity, and <span style = 'color:#FB8861FF;'>color</span>.<br>The best clarity is <span style = 'font-family:FrauncesSuperSoftWonky-MediumItalic;'>IF</span>, while the worst is <span style = 'font-family:FrauncesSuperSoftWonky-MediumItalic;'>I1</span>.<br>The best color is <span style = 'color:#FCFDBFFF;font-family:FrauncesSuperSoftWonky-MediumItalic;'>D</span>, while the worst is <span style = 'color:#51127CFF;font-family:FrauncesSuperSoftWonky-MediumItalic;'>J</span>.",
       x = "Carats",
       y = "Price",
       color = "Color:")

p
```

There is a **lot** going on here, but I'll touch on pretty much all the more obscure stuff. There is a lot of code here, but one thing to note is that the whole plot is programmatically generated with relatively little "manual" tweaking. There are a few nudges of titles (note the `hjust` and `vjust` arguments), and some manually-set hex codes for colors, but there is zero external adjusting with anything outside of R.

I'll start off talking about `theme_mcm_dark()` and building custom `ggplot2` themes, then move on to some tricks using the fantastic `ggtext` and `ggh4x` packages.

# Custom Theme

This whole plot started off with me trying to develop a cool custom `ggplot2` theme, and you can see the final version here:

```{r theme-dark}
theme_mcm_dark <- function(base_family = "FrauncesSuperSoftWonky-Light",
                           title_family = "FrauncesSuperSoftWonky-BlackItalic",
                           subtitle_family = "FrauncesSuperSoftWonky-LightItalic",
                           axis_family = "FrauncesSuperSoftWonky-Light",
                           base_color = "white",
                           primary_color = "grey10",
                           accent_color = "grey90",
                           gridlines = F,
                           facet_outlines = F,
                           base_size = 10) {
  min_theme <- theme_bw() +
    theme(
      text = element_text(
        family = base_family,
        colour = base_color,
        size = base_size
      ),
      line = element_line(
        colour = base_color,
        size = 0.2),
      legend.background = element_blank(),
      legend.key = element_blank(),
      panel.background = element_blank(),
      panel.border = element_rect(
        fill = NA,
        colour = base_color,
        size = 0.2
      ),
      strip.background = element_blank(),
      axis.line = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_blank(),
      plot.background = element_rect(fill = primary_color, color = NA),
      axis.text = element_text(
        family = base_family,
        colour = base_color,
        size = base_size
      ),
      plot.title = element_markdown(
        family = title_family,
        colour = accent_color,
        size = base_size * 1.5 ^ 2
      ),
      plot.subtitle = element_markdown(
        family = subtitle_family,
        colour = accent_color,
        size = base_size * 1.5
      ),
      axis.title = element_markdown(
        family = axis_family,
        colour = accent_color,
        size = base_size * 1.4
      ),
      strip.text = element_markdown(
        family = axis_family,
        colour = accent_color,
        size = base_size * 1.4
      ),
      legend.text = element_markdown(
        family = axis_family,
        colour = accent_color,
        size = base_size * 1.4
      ),
      legend.title = element_markdown(
        family = base_family,
        colour = base_color,
        size = base_size * 1.2
      )
    )

  if (!gridlines) {
    min_theme <- min_theme + theme(panel.grid = element_blank())
  }

  if (!facet_outlines) {
    min_theme <- min_theme + theme(panel.border = element_blank())
  }

  return(min_theme)
}
```

The goal here is not for the theme to be a be-all end-all choice, but a reasonable starting point for a lot of situations, and to provide a consistent appearance for a bunch of plots. It reduces how much I have to type out each time, but you'll see that I still ended up changing some additional stuff in `theme()` in my plot above.

Setting things like `base_size` and `base_font` is common in `ggplot2` themes, like `theme_bw()`, but I also have a couple Boolean arguments for common things I like to turn on/off: `gridlines` and `facet_outlines`. I also really like setting a bunch of default arguments as a way to reduce typing but keep flexibility for changing things like fonts if need be.

Another thing I find frustrating is changing a bunch of font sizes for different plot elements, especially when there are some general guidelines, like "the title should be bigger than the axis labels". So the way I have it set up is to set a base font size, and then scale things like `plot.title` and `axis.title` from that. You can change up the relative scalings however you like.

For inspiration on making custom themes, I'd recommend checking out Urban Institute's fantastic [style guide](https://urbaninstitute.github.io/graphics-styleguide/) and [Bob Rudis' custom themes](https://github.com/hrbrmstr/hrbrthemes).

## `element_markdown()`

You may notice that inside `theme_mcm_dark()`, there are a bunch of `element_markdown()` calls. These come from the wonderful [`ggtext` package](https://wilkelab.org/ggtext/index.html), and they replace the standard `element_text()` with text that has fuller Markdown capabilities. This is how I did stuff like use a CSS `<span>` to change the colors and fonts of text in the plot subtitle. Your text strings may get a bit bulky, but it's super powerful.

# Other Extensions

## Getting hex codes from color scale

One thing that has to be done semi-manually is pulling out hex color codes generated by `scale_color_viridis_d()` for use in the subtitle. The easiest way I know to do this is to use `ggplot_build()` to build the plot and pull out a vector of unique colors, then use the `swatchplot()` function from the [`colorspace` package](http://colorspace.r-forge.r-project.org/index.html) to view them and pick the right codes.

```{r get-colors}
g <- ggplot_build(p)

cols <- unique(g$data[[1]]$colour)

cols
colorspace::swatchplot(cols)
```

If you wanted to do this more automatically, you could probably generate a dummy plot with the correct number of color levels in order, pull out the first and last hex codes, and paste them into the `subtitle` string.

## `ggh4x`

A lot of the cool parts of this plot that would normally have to be done manually are made possible by the amazing [`ggh4x` package](https://teunbrand.github.io/ggh4x/index.html), which I cannot recommend highly enough. I'll walk through several of the functions I used, but the package can do **way** more cool stuff.

### `facet_nested()` to get variable name

I don't always love the naming options for facets. By default, the values for `cut` would just be listed on their own: `Fair`, `Good`, etc. You can use `label_both()` to generate something like `Cut: Fair`, `Cut: Good`, and so on, but that's a lot of redundant text. I far prefer to put `Cut` once, above all the facets.

The function `facet_nested()` is designed to allow nested facets, like nesting facets for cities within states, but we can use it as a hack-ish way to put `Cut` and `Clarity` above their respective facets. We make dummy variables called `cutname` and `clarityname`, then use those variables inside `facet_nested()`.

#### `strip_nested()` to bold the variable names

```{r strip-nested, eval = F}
strip = strip_nested(text_x = 
                       elem_list_text(
                         family =
                           c("FrauncesSuperSoftWonky-Medium",
                             rep(NA, 100))),
                     text_y =
                       elem_list_text(
                         family =
                           c("FrauncesSuperSoftWonky-Medium",
                             rep(NA, 100))))
```


To make the names `Cut` and `Clarity` stand out a little bit, I'd like change their text to a heavier font. However, they are just `strip` labels, treated the same as the actual facet labels like `Fair` and `Good`. We can use `strip_nested()` to set the formatting for each strip label. Since the first label is `Cut` or `Clarity`, I adjust the first strip labels, and then use `rep(NA, 100)` to say "all the other strips should be the default". This function uses vector recycling, so using 100 `NA`s is just to ensure that all the rest of the facets get an `NA`, as long as there aren't more than 100 facets...

### `guide_stringlegend()`

A relatively simple, but beautiful, function in `ggh4x` is `guide_stringlegend()`. If you have a discrete color scale, the default legend is a square of each color with a label next to it. However, `guide_stringlegend()` will just write out each label in that level's color. I had to use `title.vjust` when doing it with a horizontal legend or the title wasn't aligned with the labels.

### `facetted_pos_scales()` to remove redundant facet axis labels

One slightly complicated thing I did was remove the facet labels from all but the bottom left facet. When plotting small multiples like this, I don't think the repeated scales are all that useful, and they just clutter up the plot. However, removing them is a bit tricky.

I used the `facetted_pos_scales()` function from `ggh4x` to programmatically set the scales to be used for each facet. It requires that we set `scales = "free"` in our `facet_` function. Once we do that, we get to use nice little formulas with `~` notation, just like the `case_when()` function.

```{r faceted-scales, eval=F}
facetted_pos_scales(
    x = list(
      cut == "Fair" ~ scale_x_log10(limits = range(diamonds$carat)),
      cut != "Fair" ~ scale_x_log10(limits = range(diamonds$carat), 
                                    labels = NULL)
    ),
    y = list(
      clarity == "I1" ~ scale_y_continuous(limits = range(diamonds$price), 
                                           labels = scales::label_dollar()),
      cut != "I1" ~ scale_y_continuous(limits = range(diamonds$price), 
                                         labels = NULL)
    )
  )
```

The lefthand side of the `~` is a condition, and the righthand side is the scale to use if that condition is met. I was thinking about doing the middle facet for both `Cut` and `Clarity`, but there are an even number of levels for `Clarity` and it just didn't look great, so I opted for the bottom left facet. 

There are a couple key details to note. First, setting `labels = NULL` is used to turn off the labels for most of the facets. Second, since we set `scales = "free"`, the range of our scales will vary by row/column. Since I don't actually want the scales to be free (I only had to do it to use `facetted_pos_scales()`), I set the limits inside each scale function to the range of the x and y variables in the `diamonds` dataframe, ensuring that all the facets have the same range, just like `scales = "fixed"` would give us. Finally, inside `theme()` I need to nudge the positions of the axis titles to get them to align with this single facet.

# Conclusion

I hope this was a useful look at my process for making a custom theme and a fancier `ggplot`. I really love reading more in-depth blog posts that dig into a particular topic, but at times I find I pick up a lot of useful little tips and tricks when seeing a single thing tackled from start to finish. I hope you find this useful as well!
